The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf

### 2016-02-26 oil_prices_obes.tex
added build-out interaction effects to the model

%  oil_prices_rr_blind.tex
\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{natbib} %for citet and citep
\usepackage{syntonly}
\usepackage{esdiff} %for writing partial derivatives
\usepackage{url} %for inserting urls
\usepackage{placeins} % for limitting floats
\usepackage{comment} %for excluding figures
% \excludecomment{figure} % for printing without figures
% \syntaxonly % for quickly checking document
%set document settings

@@

\title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
% \author{Johannes Mauritzen\\
% 		Department of Industrial Economics and Technology Management\\
%         Norwegian University of Science and Technology\\
%         7491 Trondheim
%         Norway\\
%         johannes.mauritzen@iot.ntnu.no\\
%         \url{jmaurit.github.io}\\
% 		}
\date{\today}

@@

\begin{abstract}
I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.  I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 1 to 4\% for a 10 dollar per barrel increase in the real price of oil.\\
\textbf{Keywords}: Oil production,Oil Prices, Norway, Generalized Additive Model.\\
\textbf{Words}: 5970
\end{abstract}

@@

\begin{table}
\begin{center}
\begin{tabular}{l c c c }
\hline
                                             & w/out Rand. Effects & w Rand. Effects & 2-d Smooth \\
\hline
(Intercept)                                  & $0.335^{*}$    & $-4.017^{***}$ & $-574.287^{***}$ \\
                                             & $(0.146)$      & $(0.183)$      & $(51.125)$       \\
in\_place\_oil\_mill\_100sm3             & $0.337^{***}$  & $-0.014$       &                  \\
                                             & $(0.005)$      & $(0.014)$      &                  \\
cost\_index/10                            & $-0.018^{***}$ & $-0.030^{***}$ & $-0.014^{***}$   \\
                                             & $(0.003)$      & $(0.003)$      & $(0.002)$        \\
price/10                                  & $-0.006$       & $0.014$        & $0.007$          \\
                                             & $(0.009)$      & $(0.010)$      & $(0.006)$        \\
price\_l1/10                              & $0.004$        & $0.021^{*}$    & $0.015^{*}$      \\
                                             & $(0.010)$      & $(0.010)$      & $(0.006)$        \\
price\_l2/10                              & $0.037^{***}$  & $0.047^{***}$  & $0.048^{***}$    \\
                                             & $(0.009)$      & $(0.010$      & $(0.006)$        \\
price\_l3/10                              & $0.030^{***}$  & $0.025^{**}$   & $0.033^{***}$    \\
                                             & $(0.009)$      & $(0.009)$      & $(0.006)$        \\
price\_l4/10                              & $0.003$        & $0.043^{***}$  & $0.028^{***}$    \\
                                             & $(0.008)$      & $(0.008)$      & $(0.005)$        \\
price\_l5/10                              & $-0.020^{**}$  & $0.012$        & $-0.001$         \\
                                             & $(0.007)$      & $(0.007)$      & $(0.005)$        \\
price\_l6/10                              & $-0.010$       & $0.008$        & $-0.001$         \\
                                             & $(0.007)$      & $(0.007)$      & $(0.004)$        \\
price\_l7/10                              & $-0.008$       & $0.007$        & $0.004$          \\
                                             & $(0.007)$      & $(0.007)$      & $(0.004)$        \\
price\_l8/10                              & $-0.044^{***}$ & $-0.011$       & $-0.027^{***}$   \\
                                             & $(0.006)$      & $(0.007)$      & $(0.004)$        \\
EDF: s(prod\_year)                           & $6.381^{***}$  &                &                  \\
EDF: s(year)                                 & $2.768^{***}$  & $3.000^{***}$  & $3.000^{***}$    \\

EDF: s(prod\_year,re\_field)                      &                & $33.002^{***}$ &                  \\

EDF: s(prod\_year,in\_place\_oil\_mill\_sm3) &                &                & $27.445^{***}$   \\

\hline
Deviance                                     & 23062.678      & 19765.833      & 9057.204         \\
Deviance explained                           & 0.848          & 0.870          & 0.940            \\
Dispersion                                   & 5.410          & 5.965          & 2.325            \\
R$^2$                                        & 0.841          & 0.865          & 0.938            \\
GCV score                                    & 18.145         & 16.218         & 7.355            \\
Num. obs.                                    & 1088           & 1088           & 1088             \\
Num. smooth terms                            & 2              & 2              & 2                \\
\hline
\multicolumn{4}{l}{\scriptsize{$^{***}p<0.001$, $^{**}p<0.01$, $^*p<0.05$}}
\end{tabular}
\caption{Model results. Estimated standard coefficients are shown for the linear terms. Standard errors are in parenthesis. For the smoothed terms, estimated degrees of freedom are displayed, where the p-values are from F-tests of whether the terms significantly improve the fit of the model.}
\label{GAM_model_table}
\end{center}
\end{table}

@@

@@

The findings in this paper can provide some guidance on predicting the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other offshore production areas. The results suggest that lower oil prices are likely to have only modest ramifications for production in existing fields.  Anecdotal evidence however suggests that searching and exploration are being substantially curtailed. Future oil production could then be significantly affected by way of fewer new field developments.

@@

\bibliographystyle{plainnat}
\bibliography{oil_prices}

>>>>

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{natbib} %for citet and citep
\usepackage{syntonly}
\usepackage{esdiff} %for writing partial derivatives
\usepackage{url} %for inserting urls
\usepackage{placeins} % for limitting floats
\usepackage{comment} %for excluding figures
% \excludecomment{figure} % for printing without figures
% \syntaxonly % for quickly checking document
%set document settings

@@

\title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
\author{Johannes Mauritzen\\
		Department of Economics\\
        BI Norwegian Business School\\
        Trondheim Campus\\
		Trondheim, Norway\\
        johannes.mauritzen@bi.no\\
        \url{jmaurit.github.io}\\
		}
\date{\today}

@@

\begin{abstract}
I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production.  I find no significant evidence of a concurrent reaction of field production to oil prices, though a slight lagged effect is found of the magnitude of approximately 2 to 7\% for a 10 dollar per barrel increase in the real price of oil.\\
Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
Words: 5970
\end{abstract}

@@

\begin{table}
\scriptsize
\begin{center}
\begin{tabular}{l c c c c }
\hline
 & w/out Rand. Effects & w Rand. Effects & 2-d Smooth & w Build-out \\
\hline
Intercept                                  & $0.335$        & $-4.016^{***}$ & $-574.287^{***}$ & $-3.858^{***}$ \\
                                             & $(0.265)$      & $(0.297)$      & $(89.476)$       & $(0.239)$      \\
in\_place\_oil             & $0.337^{***}$  & $-0.014$       &                  & $-0.015$       \\
                                             & $(0.009)$      & $(0.022)$      &                  & $(0.024)$      \\
cost\_index                            & $-0.018^{**}$  & $-0.030^{***}$ & $-0.014^{***}$   & $-0.034^{***}$ \\
                                             & $(0.006)$      & $(0.006)$      & $(0.004)$        & $(0.006)$      \\
price                                  & $-0.006$       & $0.014$        & $0.007$          & $0.005$        \\
                                             & $(0.017)$      & $(0.016)$      & $(0.011)$        & $(0.015)$      \\
price\_l1                              & $0.004$        & $0.021$        & $0.015$          & $-0.003$       \\
                                             & $(0.017)$      & $(0.017)$      & $(0.011)$        & $(0.020)$      \\
price\_l2                              & $0.037^{*}$    & $0.047^{**}$   & $0.048^{***}$    & $0.053^{*}$    \\
                                             & $(0.017)$      & $(0.016)$      & $(0.011)$        & $(0.023)$      \\
price\_l3                              & $0.030$        & $0.025$        & $0.033^{***}$    & $0.017$        \\
                                             & $(0.016)$      & $(0.014)$      & $(0.010)$        & $(0.024)$      \\
price\_l4                              & $0.003$        & $0.043^{**}$   & $0.028^{**}$     & $0.076^{***}$  \\
                                             & $(0.015)$      & $(0.014)$      & $(0.009)$        & $(0.020)$      \\
price\_l5                              & $-0.020$       & $0.012$        & $-0.001$         &                \\
                                             & $(0.013)$      & $(0.012)$      & $(0.008)$        &                \\
price\_l6                              & $-0.010$       & $0.008$        & $-0.001$         &                \\
                                             & $(0.012)$      & $(0.011)$      & $(0.008)$        &                \\
price\_l7                              & $-0.008$       & $0.007$        & $0.004$          &                \\
                                             & $(0.012)$      & $(0.012)$      & $(0.008)$        &                \\
price\_l8                              & $-0.044^{***}$ & $-0.011$       & $-0.027^{***}$   &                \\
                                             & $(0.011)$      & $(0.011)$      & $(0.007)$        &                \\
EDF: s(prod\_year)                           & $6.381^{***}$  &                &                  &                \\
                                             & $(7.249)$      &                &                  &                \\
EDF: s(year)                                 & $2.768^{***}$  & $3.000^{***}$  & $3.000^{***}$    & $3.000^{***}$  \\
                                             & $(2.956)$      & $(3.000)$      & $(3.000)$        & $(3.000)$      \\
EDF: s(prod\_year,name)                      &                & $33.002^{***}$ &                  & $32.962^{***}$ \\
                                             &                & $(72.000)$     &                  & $(72.000)$     \\
EDF: s(prod\_year,in\_place\_oil) &                &                & $27.445^{***}$   &                \\
                                             &                &                & $(27.895)$       &                \\
price\_l1:build\_out                   &                &                &                  & $0.038$        \\
                                             &                &                &                  & $(0.024)$      \\
price\_l2:build\_out                   &                &                &                  & $-0.002$       \\
                                             &                &                &                  & $(0.029)$      \\
price\_l3:build\_out                   &                &                &                  & $0.003$        \\
                                             &                &                &                  & $(0.029)$      \\
price\_l4:build\_out                   &                &                &                  & $-0.027$       \\
                                             &                &                &                  & $(0.024)$      \\
\hline
Deviance                                     & 23062.678      & 19765.852      & 9057.204         & 19786.567      \\
Deviance explained                           & 0.848          & 0.870          & 0.940            & 0.870          \\
Dispersion                                   & 17.852         & 15.625         & 7.123            & 15.641         \\
R$^2$                                        & 0.841          & 0.865          & 0.938            & 0.865          \\
GCV score                                    & 18.145         & 16.218         & 7.355            & 16.234         \\
Num. obs.                                    & 1088           & 1088           & 1088             & 1088           \\
Num. smooth terms                            & 2              & 2              & 2                & 2              \\
\hline
\multicolumn{5}{l}{\scriptsize{$^{***}p<0.001$, $^{**}p<0.01$, $^*p<0.05$}}
\end{tabular}
\caption{Model results. Estimated standard coefficients are shown for the linear terms. Standard errors are in parenthesis. For the smoothed terms, estimated degrees of freedom are displayed, where the p-values are from F-tests of whether the terms significantly improve the fit of the model.}
\label{GAM_model_table}
\end{center}
\end{table}

@@

One potential weakness of the estimation so far is that I do not take into account the different stages of field production, where the effect of price could be expected to differ. 
In particular, in the build-out phase of a field, where production is started and then ramped up to full capacity, the field operator has an incentive to build out the field as quickly as possible as large amounts of capital have at that point been allocated and it is important to get oil production and in turn income flowing. At this stage, it is ex-ante questionable that prices would have an effect on production, as a decision to increase capacity would require a new submission to the regulators and a potentially lengthy and expensive delay. On the other hand, the draw down phase is characterized by overcapacity and in turn greater flexibility in increasing production by way of, for example, infill drilling that can both accelerate recovery and increase total recovery in a field. \footnote{I thank an anonymous referee for making this point.}

To attempt to capture this heterogeneity, I introduce an indicator variable that represents the build-out phase defined as the period between the start of production up to peak production. I add interaction effects between this indicator variable and the block of significant lagged price effects $price\_l1$ through $price\_l4$.

The main problem with introducing this heterogeneity in the model is that most field build-outs happen quickly, usually within one or two years in most small to medium sized fields. The sample to estimate the build out price terms is then relatively small - less than one third of the total observations. The small sub-sample leads to estimation problems when using the model with a 2-dimensional smooth term, so I instead extend the random effects model. I also remove the block of 5th through 8th year lagged price terms that could not be shown to increase the fit of the model.

The estimation results for this model are shown in the fourth column of table \ref{GAM_model_table}. The interaction effects, labeled $price lt:build$, are not estimated to be significantly different from zero. Given the small sample used to estimate these terms, this is not surprising. On the other hand, if we compare the main effects of the lagged price terms to the standard random effects model, we notice that the point estimates on the 2nd, 3rd and 4th lags are now subsantially larger in magnitude. In this model, these main effect terms can be interpreted as the effect of price on production in the draw-down phase. This provides some evidence that the effect of price on production is indeed strongest in this latter stage of field development.

@@

The findings in this article can provide some guidance on predicting the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other offshore production areas. The results suggest that lower oil prices are likely to have only modest ramifications for production in existing fields.  Anecdotal evidence however suggests that searching and exploration are being substantially curtailed. Future oil production could then be significantly affected by way of fewer new field developments.

Policy implications arise for both energy industry regulators and financial policy makers. Since the evidence points to a stronger lagged effect of price on production in declining fields, this would also imply that policy measures meant to stimulate increased production would likely also be more effective in mature oil producing areas compared to newer areas. This is particularily important in the Norwegian Continental Shelf, as several new fields have recently come online following a period of high oil prices. \footnote{I thank an anonymous reviewer for pointing out this implication.}

@@

\section{Acknowledgments}
I thank Klaus Mohn, Jonas Andersson, Sturla Kvamsdal, Harrison Fell, R\"ognvaldur Hannesson and Henrik Horn for valuable discussion. I have received financial support from the Norwegian Center for Sustainable Energy Studies (CenSES), Norwegian Research Council grant number 209697

\begin{spacing}{1}
\bibliographystyle{plainnat}
\bibliography{oil_prices}
\end{spacing}

### 2016-02-26 oil_prices_title.tex
added build-out interaction effects to the model

\title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
\author{Johannes Mauritzen\\
		Department of Industrial Economics and Technology Management\\
        Norwegian University of Science and Technology\\
        7491 Trondheim
        Norway\\
        johannes.mauritzen@iot.ntnu.no\\
        \url{jmaurit.github.io}\\
		}
\date{\today}

>>>>

\title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
\author{Johannes Mauritzen\\
		Department of Economics\\
        BI Norwegian Business School\\
        Trondheim, Norway\\
        johannes.mauritzen@iot.ntnu.no\\
        \url{jmaurit.github.io}\\
		}
\date{\today}
